---
title: "Investigating the exponential distribution"
subtitle: "Course Project for Statistical Inference Coursera Class"
author: "Leon Duplay"
date: "15 July 2015"
output: 
  html_document: 
    fig_height: 3
    fig_width: 6
    keep_md: yes
    theme: readable
---

## Overview

In this document, we will investigate the exponential distribution in R using the Central Limit Theorem. The probability density function of an exponential distribution is defined as f(x,l) = l*e^-lx for x>0. One key feature of the exponential distribution is that both the mean and standard deviation correspond to `1/lambda`.

In order to investigate the exponential distribution in R, we will simulate data and perform the following tasks:

1. Analyze the sample mean and compare it to the theoretical mean of the distribution according to its definition.
2. Analyze how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
3. Demonstrate that the distribution is approximately normal, focusing on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.

Below you can see an example of the exponential distribution probability density function:

```{r demo, fig.align='center'}
# Show a demo of the exponential distribution
library(ggplot2)
x <- seq(0.0,10.0,0.01)
qplot(x,dexp(x,0.2),geom="line",xlab="x", ylab="dexp(x,0.2)",main="Exponential distribution example for lambda = 0.2")
```

## Simulations

In this first part, we will simulate the needed data using `rexp(n, lambda)`. We fix the value `lambda = 0.2`, and start with `n = 40` samples. We will perform `nosims = 1000` experiments. As explained above, the theoretical mean and standard deviation of the distribution is 1/lambda, i.e. 5.

```{r init}
# Initialize parameters
set.seed(7355608)
nosims <- 1000
n <- 40
lambda <- 0.2
mean_theorical <- 1/lambda
sd_theorical <- 1/lambda

# Simulation
means_simulated <- NULL
for (i in 1: nosims) {
        means_simulated <- c(means_simulated, mean(rexp(n, lambda)))
}
```

## Sample Mean versus Theoretical Mean

According to the Central Limit Theorem, with our large number of simulations (1000), we can say: $$\frac{\mbox{Estimate} - \mbox{Mean of Estimate}}{\mbox{Std. Err. of Estimate}} = \frac{\bar X_n - \mu}{\sigma / \sqrt{n}}=\frac{\sqrt n (\bar X_n - \mu)}{\sigma} \longrightarrow N(0, 1)$$

This translates to the distribution of the sample mean $\bar X_n$ is approximately $N(\mu, \sigma^2/n)$
	- distribution is centered at the population mean
	- with standard deviation = standard error of the mean

`insert figures here`

Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

## Sample Variance versus Theoretical Variance

Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.

## Distribution

Via figures and text, explain how one can tell the distribution is approximately normal. focusing on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.



hist(runif(1000))

vs

mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(runif(40)))
hist(mns)



Now in the second portion of the class, we're going to analyze the ToothGrowth data in the R datasets package. 
Load the ToothGrowth data and perform some basic exploratory data analyses 
Provide a basic summary of the data.
Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
State your conclusions and the assumptions needed for your conclusions. 
Some criteria that you will be evaluated on
Did you  perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
Did the student perform some relevant confidence intervals and/or tests?
Were the results of the tests and/or intervals interpreted in the context of the problem correctly? 
Did the student describe the assumptions needed for their conclusions?

## Appendix

This analysis was completed with the below system:

```{r session}
sessionInfo()
```
