---
title: "Tooth Growth dataset analysis"
subtitle: "Course Project for Statistical Inference Coursera Class"
author: "Leon Duplay"
date: "16 July 2015"
output: 
  html_document: 
    fig_height: 3
    fig_width: 10
    keep_md: yes
    theme: readable
---

## Overview

In this document, we will investigate the ToothGrowth dataset by performing some basic exploratory data analysis and some statistical inference. By using confidence intervals and/or hypothesis tests to compare tooth growth by `supp` and `dose`, we will study the impact of these two variables on tooth growth and present our conclusions, including the assumptions made.

## Exploring the dataset

In this section, we will load the ToothGrowth dataset and perform some exploratory data analysis, before performing our hypothesis testing in the next section.

```{r init}
# Libraries
library(datasets)
library(ggplot2)
library(grid)
library(gridExtra)

# Load data & basic data info
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
summary(ToothGrowth)
```

The data represents 60 observations, with the length of teeth in each of 10 guinea pigs after following three dose levels of Vitamin C (0.5, 1 and 2 mg) with each of two delivery methods (orange juice OJ or ascorbic acid AC).

We'll now explore the data using some simple boxplots to get an idea if `supp` and `dose` have an impact on tooth length and in which way.

```{r explo, fig.align='center'}
# Plot length vs dose & supp
g1 <- ggplot(aes(x = dose, y = len), data = ToothGrowth) + geom_boxplot(aes(fill = dose))
g2 <- ggplot(aes(x = supp, y = len), data = ToothGrowth) + geom_boxplot(aes(fill = supp))
grid.arrange(g1, g2, ncol=2)

ggplot(aes(x = supp, y = len), data = ToothGrowth) + 
    geom_boxplot(aes(fill = supp)) + facet_wrap(~ dose)
```

In the graphs above, we can see a clear trend: the larger the dosage, the longer the tooth. However, the effectiveness of delivery method is not clear, both OJ and VC have roughly the same performance (with OJ having a slightly higher mean).

Looking at both variables together gives an interesting insight: it seems that at lower doses OJ is more effective than VC in terms of tooth length, but at high dosage (2mg), the performance is the same.

## Confidence Intervals and Hypothesis Testing

The objective of this section is to use confidence intervals and hypothesis testing to prove or disprove the null hypothesis (dosage/supplement have no impact on teeth length). To do so, we will use unpaired T-tests between factors of `dose` and `supp`.

```{r ttestsupp}
tt <- t.test(len ~ supp, paired = F, var.equal = F, data = ToothGrowth)
tt
```

In this T-test testing for supplements, while the means are different **(`r round(tt$estimate,2)`)**, we cannot disprove the null hypothesis as the probability under the null hypothesis of obtaining these means is over 5% (p value is **`r round(tt$p.value,3)`**). Therefore, we can conclude that supplement type **does NOT** have a statistically significant impact on teeth length within the given sample.

In order to use T tests on the `dose`, we must split the dataset into pairs of dosages.

```{r ttestdose}
    dose1 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
    dose2 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
    dose3 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
    t1 <- t.test(len ~ dose, paired = F, var.equal = F, data = dose1)
    t2 <- t.test(len ~ dose, paired = F, var.equal = F, data = dose2)
    t3 <- t.test(len ~ dose, paired = F, var.equal = F, data = dose3)
```

The p values for the T-tests are **`r t1$p.value`**, **`r t2$p.value`**, and **`r t3$p.value`**. Since these values are all under 0.05, we can disprove the null hypothesis and conclude that dosage **does** have a statistically significant impact on teeth length within the given sample.

## Assumptions

For all these hypothesis tests, we are assuming that our guinea ping subjects represent a population that is IID (independant and identically distributed), and that:

* The guinea pigs are repesentative of the whole population and follow a random sample.
* Tooth length shows a normal distribution
* Observations are independent of each other

## Appendix

This analysis was completed with the below system:

```{r session}
sessionInfo()
```
